\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-13 honors_thesis/journal_paper/paper.tex
Done with revisions.  I think it's ready to submit

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
20,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
effect on the value obtained.  However, increasing the swarm size from 100 to
200 makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  For this reason, previous work has recommended
the use of a swarm size of 50 for this
function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least
some cases, adding particles indefinitely will not yield an efficient
implementation.

@@

This refactoring comes at significant expense; instead of just computing
$\ofunc(\ppos_{t})$ and then $\ofunc(\ppos_{t+1})$ in the usual way, this
refactored PSO computes $\ofunc(\ppos_{t})$ and all of the
$\ofunc(\specpos_{t+1}^{c})$, eight function evaluations in this case.  The
potential value of these speculative children lies in the fact that we can
perform the computation needed for both iteration $t$ and iteration $t+1$ at
the same time.  Remember that our intent is to accelerate parallel PSO and that
we are interested in objective functions that have at least non-trivial
computational complexity.  That is, we assume that computing
$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of
PSO.

Suppose that unlimited processors were available and that the evaluation of
the objective function took one hour. Suppose also that for each particle the
computations $\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$
were sent to separate processors and completed in parallel.  Once the values
of $\ofunc(\ppos_{t})$ have been computed for all particles, the computation
needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
and to complete iteration $t+1$ is inconsequential.  In this environment all
of the function evaluations needed for two iterations of PSO could be
completed in just one hour where as the conventional implementation would
require two hours for those same two iterations.

However, we do not have an infinite number of processors at our disposal, and
performing speculative evaluation significantly increases the number of
function evaluations needed in PSO.  We show in the rest of this paper that in
many, though not all, instances the extra performance achieved by being able to
do more iterations at a time outweighs the benefit that would have been gained
by merely increasing the number of particles in the swarm.

>>>>

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is slightly larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

To see the value of this refactoring, suppose that 1000 processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of 100 particles, 900 of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
rest of this paper that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.
